---
title: "Project_1_456"
author: "Anthony Yasan, Preston O'Connor, Khoa Dao"
date: "2025-02-20"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

We are modeling the linear regression of the Dependent Income, Independent Age in our model

# Introduction

## Installing the R-packages

```{r}
# remove comments out these blocks to install the R packages that are being used
#install.packages("ipumsr") # for the data set
#install.packages("dplyr") # for the data set
#install.packages("caTools") # use this for the set seed of the training set
#install.packages("ggplot2") 

# Code to implement the R packages
library(ipumsr)
library(dplyr)
library(ggplot2) # visial displays of the Boxplot, and Q-Q plots
library(caTools)
```

# Data description

The dataset is sourced from IPUMS USA, which provides microdata extracted from the U.S.
Census and American Community Survey (ACS).
It contains demographic and economic data at both household and individual levels.

The dataset consists of 3,405,809 rows and 15 columns.
Each observation represents a household.

Here are the key variables included in the dataset:\
- YEAR: Census year (e.g., 2023).\
- SAMPLE: IPUMS sample identifier.\
- SERIAL: Unique household serial number.\
- CBSERIAL: Original Census Bureau household serial number.\
- HHWT: Household weight for proper representation.\
- CLUSTER: Household cluster for variance estimation.\
- STRATA: Household strata for variance estimation.\
- GQ: Group quarters status (e.g., household, institution).\
- HHINCOME: Total household income for all members over 15 years old.
- PERNUM: Person number within the household.\
- PERWT: Person weight for population estimates.\
- SEX: Gender classification (Male/Female).\
- AGE: Individual’s age in years.\
- RACE: General race classification.\
- RACED: Detailed race classification.

## Information about the Data set

```{r}
ddi <- read_ipums_ddi("usa_00001.xml")
data <- read_ipums_micro(ddi)
dim(data)
summary(data)
```

## Table of Data

```{r}
#View(data)
#The Code struggles to run the data set with 2 million points is two extensive to run
set.seed(11)

s <- sample(1:nrow(data), size = 200000)
data <- data[s, ]
dim(data)

```

## Data Cleaning and Outlier Removal

```{r}
# select the age and the Total Household income as the main columns of interest, then filter based of 18 <= age <= 65 (Thats the only age bracket that is currently working and has other factors effecting income). Also, there is a negative value in household income which doesn't make sense at all, so we filter it out. 

# ask if the filter crashes out after a certain amount on the computer and if we need to shrink the train size

data <- data %>%
  select(AGE, HHINCOME) %>%
  mutate(HHINCOME = as.numeric(HHINCOME), AGE = as.numeric(AGE)) %>%
  filter(!is.na(HHINCOME), !is.na(AGE)) %>%
  filter(between(AGE, 18, 65)) %>%
  filter(HHINCOME > 0)

dim(data)  #if you want to view the two filtered columns

IQR_of_AGE <- IQR(data$AGE)
IQR_of_HHINCOME <- IQR(data$HHINCOME)

# calculating the upper and lower bounds of both of the data sets to filter the data
AGE_lower <- quantile(data$AGE, 0.25) - 1.5 * IQR_of_AGE
AGE_upper <- quantile(data$AGE, 0.75) + 1.5 * IQR_of_AGE

HHINCOME_lower <- quantile(data$HHINCOME, 0.25) - 1.5 * IQR_of_HHINCOME
HHINCOME_upper <- quantile(data$HHINCOME, 0.75) + 1.5 * IQR_of_HHINCOME

#continue to filter any of the outliers that are presents in the data set
filtered_data <- data %>%
  filter(between(AGE, AGE_lower, AGE_upper)) %>%
  filter(between(HHINCOME, HHINCOME_lower, HHINCOME_upper))

```

## Original Histogram

### Age
```{r}
ggplot(data, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency") +
  theme_minimal()
```
### Household Income
```{r}
ggplot(data, aes(x = HHINCOME)) +
  geom_histogram(binwidth = 10000, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Household Income", x = "Household Income", y = "Frequency") +
  theme_minimal()
```

## Filtered Histogram

### AGE
```{r}
ggplot(filtered_data, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency") +
  theme_minimal()
```
### Household Income
```{r}
ggplot(filtered_data, aes(x = HHINCOME)) +
  geom_histogram(binwidth = 10000, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Household Income", x = "Household Income", y = "Frequency") +
  theme_minimal()
```

## Orignal Box Plots

### AGE
```{R}
boxplot(data$AGE, main = "Boxplot of Age", col = "lightblue", ylab = "Age")
```

### Household Income
```{R}
boxplot(data$HHINCOME, main = "Boxplot of Household Income", col = "lightgreen", ylab = "Household Income")
```

## Filtered Box Plots

### AGE
```{R}
boxplot(filtered_data$AGE, main = "Boxplot of Filtered Ages", col = "lightblue", ylab = "Age")
```
### Household Income
```{R}
boxplot(filtered_data$HHINCOME, main = "Boxplot of Filtered Household Income", col = "lightgreen", ylab = "Household Income")
```

#Analysis
```{r}
# modifying data into a training set and a testing set
set.seed(1)
# ask about a good metric for the split of the data
split <- sample.split(filtered_data$HHINCOME, SplitRatio = 0.98)
train_set <- subset(filtered_data, split == TRUE)
test_set <- subset(filtered_data, split == FALSE)

#sized of the sets
dim(train_set)
dim(test_set)

#model from the training data
linear_model <- lm(HHINCOME ~ AGE, data = train_set)

# Prediceted values on the test set
test_set$predicted_HHI <- predict(linear_model, newdata = test_set)

# calculate residuals for the test set
test_set$residuals <- test_set$HHINCOME - test_set$predicted_HHI


#implement the diagonal plot 

# implement the various forms of analysis to show and explain what is going on in the data set
```

## Implementing the Plots

### Risidual vs. Fitted Values Plot
```{r}
ggplot(test_set, aes(x = predicted_HHI, y = residuals)) +
  geom_point(alpha = 0.5, color = 'black') +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

###Normal Q-Q Plot
```{r}
ggplot(test_set, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of Residuals") +
  theme_minimal()
```

### Linear Regression with Testing Data
```{r}
ggplot(test_set, aes(x = AGE, y = HHINCOME)) +
  geom_point(alpha = 0.3, color = "red") +  # Scatter plot of data points
  geom_smooth(method = "lm", color = "Navy", se = FALSE) +  # Regression line
  labs(title = "Linear Regression: House Hold Income vs Age",
       x = "Age of Person",
       y = "Total House Hold Income") +
  scale_y_continuous(labels = scales::comma)
  theme_minimal()
```

### Summary of the Simple Linear Regression Model
```{r}
summary(linear_model)
```

# Model Evaluation and Prediction

# Conclusion and Summary

# Reference
